Gamma-Ray Burst Afterglows In The Multi-Messenger Era: Numerical Models and Closure Relations

### 2020-05-13 refereeResponse.txt

We thank the referee for their thoughtful report and comments. They highlighted several areas where the text was not clear, the revised manuscript is much improved.  Our responses to the referee's comments are below, with the associated changes to the text.

1. My main concern is the Lorentz factor profile which is currently modeled. Right now, it is assumed that the fluid elements at higher latitudes would contribute to the observed emission along the line of sight when their Lorentz factor becomes the same as that along the line of sight. This is not necessarily the case in reality. In my mind, different fluid elements along different polar angles may evolve differently, and the observed emission is roughly from a smaller 1/\gamma half cone near the polar region and a larger 1/\gamma half cone from the wing region. I suppose when such a detailed modeling is included, the lightcurves would be different for different \Gamma(\theta) profiles even if the E(\theta) and \theta_{obs} are the same. The authors should emphasize this issue throughout the paper (including in the abstract and conclusions), especially when closure relations are presented, so that future users can keep this caveat in mind when they apply the relations.

--

The referee is correct: fluid elements at different latitudes evolve differently, and the emitting region is asymmetric due to the differing Lorentz factors (as well as time-of-flights for the emitted photons) along the latitudinal direction.  However, these effects are already accounted for in our numerical model (afterglowpy).  Each annular section of the jet evolves independently, with its own Lorentz factor dependent on its initial energy.  We then integrate over all sections, taking into account doppler factors and time-of-flight, to arrive at a final flux.

We do not assume "fluid elements at higher latitudes would contribute to the observed emission along the line of sight when their Lorentz factor becomes the same as that along the line of sight".  In our numerical model we integrate over the entire jet surface, relativistic doppler factors naturally and appropriately reduce the contribution of sections which are beamed away from the observer.  In our analytic modeling, we assume regions whose beaming cone includes the observer contribute to the flux, but the flux is dominated by the brightest of these regions.  The brightest region will be that with the highest Lorentz factor, which will be located closest to the jet core.  This will be the case so long as E(theta) is a decreasing function of theta.

Once each section of the jet begins decelerating, the initial Lorentz factor profile gamma(theta) becomes less important.  An ultra relativistic, adiabatic, decelerating blast wave obeys E \propto R^3 \gamma^2, hence \gamma \propto E^{1/2} R{-3/2}.  Given an initial profile E(\theta), the profile \gamma(\theta) in the deceleration regime is fixed.  This is true for trans-relativistic blast waves too, although the particular form of the relation \gamma ~ E,R relation will change.  The initial \gamma(\theta) plays an important role in the early evolution of the blast wave and its observed light curve, but is forgotten as more of the jet begins deceleration.

We have added remarks clarifying the role of the Lorentz factor profile and its treatment throughout the text.

-- added "decelerating" to the description of the jets under consideration in the abstract.

-- added Section 2.1: To clarify the role of the Lorentz factor profile, and justify our focus on the energy profile) we expanded the last paragraph of section 2 into a new section 2.1.

-- moved start of Section 3.1 two paragraphs down, appended "Top-Hat Case" to the name

-- added new Section 3.2 detailing how top-hat sections are integrated into a structured jet light curve, emphasizing the non-trivial profile of shock radius and velocity (e.g. Lorentz factor)

-- added two sentences in the "Coasting" paragraph of Section 4.1 noting that we do not model the initial Lorentz factor profile and so may miss some phenomenology in the transition to the later phases of evolution.

-- expanded the explanation of the "structured" phase in Section 4.1 to emphasize the role of the Lorentz factor and the asymmetric shape of the emitting region.

-- Amended the captions of Table 1 and Figure 3 to denote these apply only when the jet is decelerating.

-- Added a sentence to the summary emphasizing the results are only appropriate for decelerating jets.

2. I understand that the authors introduced s_\Omega to partially account for the uncertainty of E(\theta) effect and probably \Gamma(\theta) effect as well. However, I do not understand why the typical value is s_\Omega = 1. I understand that the opening angle scales as 1/\gamma_*. However, the solid angle should carry the dimension of (1/\gamma_*)^2, so that s_\Omega should be approximately 2 to me.

--

s_\Omega is certainly equal to 2 in a standard afterglow when the emitting surface has a (mostly) uniform Lorentz factor.  In the structured phase, we can think of the solid angle as being the product of two parts:  \Delta Phi and \Delta \Theta.  In the azimuthal direction (a line of constant theta), the Lorentz factor is uniform and so \Delta \Phi \propto 1 / \gamma_*  as expected.  In the polar direction (a line of constant phi), however, the situation is not as simple.  We consider a point theta_* on the jet which is the centroid of emission.  If dE/d\theta (or equivalently d\gamma/d\theta) is sufficiently large, then material at theta_* - 1/gamma_* (towards the core) will have substantially larger Lorentz factor and be beamed away from the observer.  Material at \theta_* + 1/gamma_* (away from the core) will have substantially lower Lorentz factor and be subdominant to the emission.  Therefore \Delta \theta can not equal 1/\gamma_*.  In this case it is dE/d\theta which controls the visible solid angle in the polar direction, not \gamma_*.  The solid angle depends on the Lorentz factor like 1/gamma_* and s_\Omega = 1.

In detail s_\Omega will smoothly transition between 2 (in the standard afterglow) 1 (during a structured phase) and 0 (after the jet break).  The detailed evolution depends on the particular energy profile and parameters, but in our explorations (as shown in Figs 3 and 7) s_\Omega = 1 is a good approximation for the bulk behaviour during the structured phase.

Our expanded discussion of the structured phase in Section 4.1 goes into more detail about the shape of the emitting region and why s_\Omega can equal 1.

3. When numerical models reach accuracy better than 10^{-4}, one should avoid introducing physical inaccuracy larger than 10^{-4}. I don't know what is the inaccuracy introduced by the crude treatment of synchrotron radiation, but another source of inaccuracy could originate from the equation of state. According to Eqs.(5) and (6), the adiabatic index \hat\gamma is adopted as 3/4. This is the extreme value for a relativistic fluid. The approximation may not be good for large viewing angles at late times (as is the case for modeling GRB 170817A). If one writes \hat\gamma = (4 \bar\gamma _ 1) / 3 \bar\gamma (\bar\gamma is the average Lorentz factor of shocked protons), one can naturally transit from the relativistic regime (4/3) to the non-relativistic regime (5/3). The formulae also appear more elegant. I suggest the authors adopt such a more precise treatment to match the 10^{-4} numerical accuracy. See e.g.
https://ui.adsabs.harvard.edu/abs/2003ApJ...591.1075K/abstract
https://ui.adsabs.harvard.edu/abs/2011ApJ...733...86U/abstract

--

The referee is mistaken about our equation of state. Equations 5 and 6, as well as all of afterglowpy, are computed using the TM equation of state of Mignone, Plewa, and Bodo 2005 which smoothly interpolates between ultra-relativistic (4/3) and non-relativistic (5/3) adiabatic indices.  This is the same EoS used by the referees ' suggested paper Uhm 2011.

We note for a structured jet the accuracy of our numerical scheme is closer to 10^{-2} than 10^{-4}, as stated at the end of Section 3.3. We do not claim the overall accuracy of the method is at either of these levels, merely the numerical accuracy of the physics we include.  The comparison to BoxFit in section 3.4 is a better (though still not perfect, given BoxFit's similar synchrotron model) measure of our overall accuracy, which is seen to be nearer ~ few * 10^{-1}.

We amended the manuscript to make the choice of equation of state, and its reasons for selection, much more clear.

-- Added several sentences to the second paragraph of section 3.1, providing more explicit details about the equation of state, incorporating the latter reference suggested by the referee, and more references found within.

4. I don't understand why the authors introduce a truncated angle \theta_w for Gaussian jets. It is essentially impossible in reality. Also for GRB 170817A, it seems that the typical viewing angle is greater than \theta_w. If this is the case, one should not expect to have relativistic ejecta along the line of sight in the early phase. This is apparently in contradiction with the prompt emission data, which require a relativistic outflow along the line of sight.

--

The truncation angle \theta_w serves two purposes.  Firstly is to do with numerical efficiency: high latitude wings of the jet have a negligible contribution to the flux (except at very early times) compared to the jet core but require a similar amount of computational time.  Excising these parts of the jet allows for quicker evaluations and the possibility of doing MCMC-type analyses which require many light curves be generated.  Secondly there are simulations, like the Aloy 2005 or Xie/Margutti 2018 works, which do show a very steep energy profile at some large angle. Truncating at theta_w is more extreme than what these simulations show, but is a rough approximation for a tightly collimated jet with structured interior.  Ultimately we find no harm in including theta_w as an optional parameter, any particular application may set it sufficiently high as to be ignorable if it wishes.

We do not yet know the precise origin of the prompt emission in GRB 170817A. It need not have originated from material directly in our line of sight. Both Matsumoto+ 2019 (https://ui.adsabs.harvard.edu/abs/2019MNRAS.483.1247M/abstract) and Ioka+ 2019 (https://ui.adsabs.harvard.edu/abs/2019MNRAS.487.4884I/abstract) consider the prompt emission in a structured jet scenario and find solutions where the prompt emission is emitted several degrees off the line-of-sight from material with a moderate Lorentz factor.  This is analogous to the structured afterglow case where the emission can be dominated by material at intermediate angles between the line of sight and jet axis.

In our fit to the 170817 afterglow we keep theta_w as a free parameter to remain agnostic about its true value.  As seen in Table 3 and Figure 12, the fit really does not constrain theta_w too much. The median value (0.47) is in fact greater than the median viewing angle (0.40), but there are numerous solutions where the hierarchy is reversed. We agree that taking the prompt emission into account would likely require a larger theta_w than the afterglow alone, but this would depend on the details of the prompt emission model and is out of scope for the current work.

This is an important point which was not discussed in the original manuscript, we have added discussion about theta_w and the prompt emission to Section 5.

-- Added a paragraph concerning theta_w and the choice of permissive prior in light of the prompt emission to Section 5.

5. Why didn't the authors consider synchrotron self-absorption? For GRB 170817A, the current modeling is adequate, but maybe in the future one needs to consider \nu_a when modeling off-axis jets.

--

Synchrotron self-absorption is an important process and certainly due a similar analysis in light of off-axis viewing and jet structure.  A proper treatment of self-absorption is complicated, particularly in off-axis viewing when much of the emitting region is at an oblique angle to the observer and near portions of the jet may shadow distant portions.  This adds significant difficulty to both the numeric and analytic understanding of the afterglow in these regimes.

We saw our development of the theory at frequencies above \nu_a to be sufficient for this particular manuscript. Incorporating frequencies below \nu_a will require more care, which we hope to present in future work.

6. The description of the early history of the structured jet model may be revised: Lipunov et al. was talking about "standard energy" and a three-component jet model. Rossi et al. proposed b=2 power law "universal" model. Zhang & Meszaros proposed "quasi-universal" (jet structures can differ somewhat among bursts) models that include both power laws and Gaussian. Nakar et al's test was strictly for the universal b=2 power law structured jet model. The model does not work even without their test: the observed E_iso - \theta_j relation is not a strict E_iso \propto \theta_j^{-2} relation (and the beaming-corrected energy is not strictly constant). In fact, the quasi-universal models, especially the quasi-universal Gaussian jet model works well for long GRBs and probably for short GRBs as well. See early tests of these models for long GRBs:
https://ui.adsabs.harvard.edu/abs/2004ApJ...601L.119Z/abstract
https://ui.adsabs.harvard.edu/abs/2005ApJ...621..875D/abstract
These facts may be clarified in the introduction.

--

We thank the referee for clearing up the history of the subject, and have amended the introduction as necessary.

-- Expanded discussion and added references concerning history of structured jets in paragraphs 5 and 6 of Section 1.

7. Figure 1 presented some structured jet models from numerical simulations. Figure 4 of this paper
https://ui.adsabs.harvard.edu/abs/2019ApJ...877L..40G/abstract
provide more structures. In particular, the shape of E(\theta) depends on the waiting time before the jet is launched. For a delayed launch, one has essentially a two-component jet structure (jet + cocoon). For a prompt launch (negligible delay), the jet penetrates through the ejecta quickly and develop a structured jet outside the ejecta. The jet structure is one-component, similar to the Gaussian profile promoted in this paper. This may be discussed in the paper.

-- Added a sentence to the end of the fourth-to-last paragraph of Section 5 explaining the origin of the constraining power on theta_obs and theta_c.

-- Added a paragraph to section 6.1 discussing the role of gravitational wave data in constraining theta_obs and theta_c.

### 2020-05-13 structuredJets.tex
working on ref report

\begin{abstract}
	Gamma-ray bursts (GRBs) associated with gravitational wave events are, and will likely continue to be, viewed at a larger inclination than GRBs without gravitational wave detections.  As demonstrated by the afterglow of \gwbns{}, this requires an extension of the common GRB afterglow models which typically assume emission from an on-axis top hat jet.  We present a characterization of the afterglows arising from structured jets, providing a framework covering both successful and choked jets.  We compute new closure relations for structured jets and compare with the established relations for energy injection and refreshed shock models.  The temporal slope before the jet break is found to be a simple function of the ratio between the viewing angle and effective opening angle of the jet.  A numerical model to calculate synthetic light curves and spectra is publicly available as the open source \python{} package \afterglowpy{}.

@@

Jet structure and viewing angle have a long history as potential explanations for temporal behaviors and breaks in GRB afterglows.  \citet{Meszaros:1998aa} first considered anisotropic models and noted they may present as ``orphan afterglows'' without a prompt GRB signal if viewed sufficiently off-axis.  Detailed theoretical calculations of structured jet afterglow light curves concluded several facts: on-axis structured light curves resemble those from top hat jets, off-axis structured light curves show an achromatic break at time $t_b\propto \thobs^{8/3}$ (where $\thobs$ is the viewing angle), and the pre-break slope can depend on the viewing angle and particular structure model \citep{Rossi:2002aa, Dalal:2002aa, Granot:2002aa, Panaitescu:2003aa, Kumar:2003aa, Granot:2003aa, Salmonson:2003aa, Rossi:2004aa}. The ``Universal Structured Jet'' was a structured jet with isotropic-equivalent energy $E \propto \theta^{-2}$ proposed to explain the diversity of jet-break times as a viewing angle effect  \citep{Lipunov:2001aa, Zhang:2002aa}, although this was ultimately unsuccessful \citep{Nakar:2004aa}.

The jet break in the structured jet afterglow light curve can exhibit a larger jump between pre- and post-break slopes than in top hat models, and has been invoked to explain some observed jet breaks which do not easily fit the standard closure relations \citep{Panaitescu:2005aa, Panaitescu:2005ab}, although other dynamical and spectral processes can have similar effects \citep{Piro:2005aa, Corsi:2006aa}.  GRB prompt emission viewed significantly off-axis is one of the possible origins for X-ray flashes \citep{Ioka:2001aa, Yamazaki:2002aa, Yamazaki:2003aa, Peng:2005aa, DAlessio:2006aa}.  More recently, small non-zero viewing angles have been measured in a subset of the \swiftXRT{} afterglow sample \citep{Ryan:2015aa, Zhang:2015aa, Troja:2016aa}.  These studies focused on observers positioned within the core opening angle of the jet or relied on closed numerical codes to calculate light curves for misaligned viewers.

@@

% Explicitly say we were the first to find X-ray, predict steep post-break slope.

@@

The Lorentz factor profile $\gamma(\theta)$ of a jet may also obtain angular structure through its launching and evolution.  In this work we focus on the consequences of non-trivial $E(\theta)$, as the bulk of successful observations are expected to occur when the jet is in the deceleration regime and the jet is locked to $\gamma(\theta) \propto E(\theta)^{1/2}$.  We leave the specific consequences of differing $\gamma(\theta)$ profiles, which may affect emission at early times, to future work.

@@

\subsection{The Single Shell Approximation}\label{subsec:algo}

@@

To accommodate an initial structure profile $E(\theta)$ we consider the integrand of Equation \eqref{eq:flux} as a function of the polar angle $\theta$.  We assume each constant-$\theta$ annulus evolves independently, as an equivalent top hat of initial width $\theta_j = \theta$.  This is a very good approximation when transverse velocities are low: when the blast wave is ultra-relativistic and has not begun to spread and when the blast wave is non-relativistic and the spreading has ceased \citep{van-Eerten:2010aa}.  Once jet spreading begins in earnest the errors are larger, and this approach can be best viewed as an interpolation between the correct ultra-relativistic and non-relativistic limits \citep{van-Eerten:2010aa}.

To compute the light curve of a top-hat jet of initial width $\theta_0$ we first must calculate the time evolution of the blast wave. In the single-shell approximation we treat the ejecta mass, contact discontinuity, and forward shock as a single unit propagating through a cold ambient medium with constant rest-mass density $\rho_0 = \Mp n_0$.  We utilize the ``TM'' trans-relativistic equation of state to describe the fluid consistently throughout its evolution \citep{Mignone:2005aa}.  The forward shock is at radius $R$ from the explosion, and the fluid behind the shock has dimensionless four-velocity $u$ and Lorentz factor $\gamma$.  The shock jump conditions can be used to determine the shock speed:
\begin{equation}
	\dot{R} = \frac{4 u \gamma}{4 u^2 +3}c\ . \label{eq:Rdot}
\end{equation}
Here the time derivative $\dot{R}$ is taken with respect to elapsed time in the bursters' frame $t$.

@@

Given an expression for the total blast wave energy $E$ in terms of $t$, $R$ and $u$, Equations \eqref{eq:Rdot}, \eqref{eq:E}, and \eqref{eq:thetadot} are sufficient to write a system of ordinary differential equations (ODEs) in $R(t)$, $u(t)$, and $\theta_j(t)$. A standard adiabatic blast wave maintains constant $E$, while energy injection and refreshed shock models may add dependence on $t$ or $u$. The general form of the shock evolution equations are then:
\begin{align}
	\dot{R} &= \frac{4 u \gamma}{4 u^2 +3}c  \label{eq:ODE1}\\
	\dot{\theta}_j &= \frac{1}{2\gamma}\sqrt{\frac{2u^2+3}{4u^2+3}} \frac{\dot{R}}{R}  \label{eq:ODE2} \\
	\dot{u} &= -\frac{E_{sw}(4u^2+3)\beta^2 (3\dot{R}/R + \cot (\theta_j/2) \dot{\theta}_j) - \partial_t E}{2 E_{sw} u (4u^4 + 8u^2 + 3)\gamma^{-4} + \beta \Mej c^2 - \partial_u E}  \label{eq:ODE3} \\
	E_{sw} &\equiv \frac{4\pi}{9} \rho_0 c^2 R^3 f_\Omega\ . \nonumber
\end{align}
Equations \eqref{eq:ODE1}-\eqref{eq:ODE3} define a three dimensional system of ODEs in the variables $(R, u, \theta_j)$ which may be solved numerically or, in certain limits, analytically from appropriate initial conditions.  Once the shock evolution ($R(t)$, $u(t)$, $\theta_j(t)$) is known, the flux is given by the integral:
\begin{align}
	F_\nu(\tobs, \nuobs) &= \frac{1+z}{4\pi \dL^2} \int \! \dd \Omega\ R^2\ \Delta R\  \delta^2\ \epsilon'_{\nu'} \ , \label{eq:flux2}
\end{align}
where $\Delta R$ is the effective shock width contributing to the emission \citep{van-Eerten:2010aa}:
\begin{equation}
	\Delta R = \frac{1}{1-\mu \dot{R}} \frac{R}{12\gamma^2} \ . \label{eq:dr}
\end{equation}
 The integrand is evaluated at a constant observer time $\tobs$ and observer frequency $\nuobs$, related to $t$ and $\nu'$ by:
\begin{align}
	\tobs &= (1+z) \left(t - \mu(\theta, \phi) R(t)/c\right)\ , \label{eq:tobs} \\
	\nuobs &= (1+z)^{-1} \delta \nu' , \label{eq:nuobs}
\end{align}

@@

@@

The numerical accuracy of a top-hat light curve with this integration scheme is typically better than $10^{-4}$.  The structured jet calculation splits the integration domain into {\tt latRes} disjoint annuli, each evolved as an independent top-hat and their emission summed.  By default {\tt latRes} is chosen such that there will be 5 zones per $\thC$-sized interval of $\theta$.   The choice of  {\tt latRes} has the largest impact on the code performance and accuracy.  We find the choice $\mathtt{latRes} = 5$ gives sufficiently quick performance at acceptable errors, typically on the order of $10^{-2}$.

@@

Due to relativistic beaming the flux at any given time is dominated by a small patch of blast wave surface.  The angular coordinates of this patch are $\theta = \theta_*$ and $\phi=0$, it has angular size $\Delta \Omega$, Lorentz factor $\gamma_*$, and emits a specific intensity $I_\nu^*$. We can write the received flux as:
\begin{equation}
	F_\nu = I_\nu^*\ \Delta \Omega\ .
\end{equation}

@@

\emph{Structured} --- If the viewer lies outside the core of the blast wave, $\thobs > \thC$, non-trivial angular structure can have a significant effect on the light curve.  Material from the energetic regions progressively closer to the jet axis continuously comes into view, becoming the new dominant source of emission, and moving the centroid with time: $\theta_* \to 0$ as $\tobs$ increases.  The centroid obeys $\gamma^{-1}(\theta_*) \sim |\thobs-\theta_*|$.  The angular size of the visible patch in the azimuthal direction is still controlled by beaming, but in the polar direction it is controlled by the steep gradient in $E(\theta)$, resulting in $\som \approx 1$.  The temporal evolution of the light curve depends on both the specific angular structure $E(\theta)$ and the viewing angle $\thobs$. The specific closure relations are given in Table \ref{tab:slopes} and discussed in Section \ref{subsec:structure}.

@@

>>>>

\begin{abstract}
	Gamma-ray bursts (GRBs) associated with gravitational wave events are, and will likely continue to be, viewed at a larger inclination than GRBs without gravitational wave detections.  As demonstrated by the afterglow of \gwbns{}, this requires an extension of the common GRB afterglow models which typically assume emission from an on-axis top hat jet.  We present a characterization of the afterglows arising from structured jets, providing a framework covering both successful and choked jets.  We compute new closure relations for {\bf decelerating} structured jets and compare with the established relations for energy injection and refreshed shock models.  The temporal slope before the jet break is found to be a simple function of the ratio between the viewing angle and effective opening angle of the jet.  A numerical model to calculate synthetic light curves and spectra is publicly available as the open source \python{} package \afterglowpy{}.

@@

Jet structure and viewing angle have a long history as potential explanations for temporal behaviors and breaks in GRB afterglows.  \citet{Meszaros:1998aa} first considered anisotropic models and noted they may present as ``orphan afterglows'' without a prompt GRB signal if viewed sufficiently off-axis.  Detailed theoretical calculations of structured jet afterglow light curves concluded several facts: on-axis structured light curves resemble those from top hat jets, off-axis structured light curves show an achromatic break at time $t_b\propto \thobs^{8/3}$ (where $\thobs$ is the viewing angle), and the pre-break slope can depend on the viewing angle and particular structure model \citep{Rossi:2002aa, Dalal:2002aa, Granot:2002aa, Panaitescu:2003aa, Kumar:2003aa, Granot:2003aa, Salmonson:2003aa, Rossi:2004aa}.

{\bf A three component jet model viewed at a range of angles was proposed to unify GRBs to a standard energy \citep{Lipunov:2001aa}.  The ``Universal Structured Jet'' was a structured jet with isotropic-equivalent energy $E \propto \theta^{-2}$ proposed to give GRBs a standard energy and explain the diversity of jet-break times as a viewing angle effect  \citep{Rossi:2002aa}, although this was ultimately unsuccessful \citep{Nakar:2004aa}.  Quasi-universal structured jets with both power-law \citep{Zhang:2002aa} and Gaussian \citep{Zhang:2004aa} profiles have also been proposed.  The Gaussian models are consistent a number of observed correlations in the GRB population and provide a possible origin for X-ray flashes \citep{Dai:2005aa}.}

The jet break in the structured jet afterglow light curve can exhibit a larger jump between pre- and post-break slopes than in top hat models, and has been invoked to explain some observed jet breaks which do not easily fit the standard closure relations \citep{Panaitescu:2005aa, Panaitescu:2005ab}, although other dynamical and spectral processes can have similar effects \citep{Piro:2005aa, Corsi:2006aa}.  {\bf GRB prompt emission viewed significantly off-axis is one of the possible origins for X-ray flashes \citep{Ioka:2001aa, Yamazaki:2002aa, Yamazaki:2003aa, Zhang:2004aa, Dai:2005aa, Peng:2005aa, DAlessio:2006aa}.  }More recently, small non-zero viewing angles have been measured in a subset of the \swiftXRT{} afterglow sample \citep{Ryan:2015aa, Zhang:2015aa, Troja:2016aa}.  These studies focused on observers positioned within the core opening angle of the jet or relied on closed numerical codes to calculate light curves for misaligned viewers.

@@

@@

\subsection{The Lorentz Factor Profile $\gamma(\theta)$}

{\bf 
The Lorentz factor profile $\gamma(\theta)$ of a jet will also obtain angular structure through the jet's launching and evolution.  Two jets with identical energy profiles $E(\theta)$ may very well acquire different Lorentz factor profiles and display different afterglow light curves.  However, these differences are restricted to the early phases of the jet's evolution.

As shown in Section \ref{subsec:algo} and Appendix \ref{app:derive1}, once a jet sweeps up enough circumburst material to begin deceleration the Lorentz factor evolves according to $\gamma(t; \theta) \propto \sqrt{E(\theta) / n_0} t^{-3/2}$, where $t$ is time in the burster's frame and $n_0$ is the (uniform) circumburst density.   While the jet is coasting, before deceleration, two jets may have identical energy profiles but differing Lorentz factor distributions.  Once deceleration begins the jet forgets its initial Lorentz factor profile $\gamma(\theta)$ and approaches one determined exclusively by $E(\theta)$.  The bulk of GRB afterglow observations, including those of \grbbns{}, are carried out after deceleration has begun.

 In this work we focus on the consequences of non-trivial energy profiles $E(\theta)$ for jets in the deceleration regime.  The energy profile is also a useful quantity to focus on as, due to conservation of energy and unlike the Lorentz factor, $E(\theta)$ is constant through the coasting and deceleration phases until the jet begins to spread at late times.  We leave the specific consequences of generic $\gamma(\theta)$ profiles, which will affect the early time light curve (see \cite{Beniamini:2020ab} for the case of power law profiles), to future work. 
}

@@

@@

To accommodate an initial structure profile $E(\theta)$ we consider the integrand of Equation \eqref{eq:flux} as a function of the polar angle $\theta$.  We assume each constant-$\theta$ annulus evolves independently, as an equivalent top hat of initial width $\theta_j = \theta$.  This is a very good approximation when transverse velocities are low: when the blast wave is ultra-relativistic and has not begun to spread and when the blast wave is non-relativistic and the spreading has ceased \citep{van-Eerten:2010aa}.  Once jet spreading begins in earnest the errors are larger, and this approach can be best viewed as an interpolation between the correct ultra-relativistic and non-relativistic limits \citep{van-Eerten:2010aa}.
{\bf 
\subsection{The Single Shell Approximation -- Top-Hat Case}\label{subsec:algo}
}

To compute the light curve of a top-hat jet of initial width $\theta_0$ we first must calculate the time evolution of the blast wave. In the single-shell approximation we treat the ejecta mass, contact discontinuity, and forward shock as a single unit propagating through a cold ambient medium with constant rest-mass density $\rho_0 = \Mp n_0$.

{\bf An accurate equation of state is essential in trans-relativistic calculations \citep{Mignone:2007aa}.  We utilize the ``TM'' trans-relativistic equation of state to describe the fluid consistently throughout its evolution \citep{Mathews:1971aa, Mignone:2005aa}.  This equation of state smoothly interpolates from a relativistic gas with adiabatic index $\hat \Gamma = 4/3$ at high temperature to a non-relativistic $\hat \Gamma = 5/3$ gas at low temperature, and has been successfully used for many GRB studies \citep[e.g.][]{Uhm:2011aa, van-Eerten:2012aa}. } The forward shock is at radius $R$ from the explosion, and the fluid behind the shock has dimensionless four-velocity $u$ and Lorentz factor $\gamma$.  The shock jump conditions can be used to determine the shock speed:
\begin{equation}
	\dot{R} = \frac{4 u \gamma}{4 u^2 +3}c\ . \label{eq:Rdot}
\end{equation}
Here the time derivative $\dot{R}$ is taken with respect to elapsed time in the bursters' frame $t$.

@@

Given an expression for the total blast wave energy $E$ in terms of $t$, $R$ and $u$, Equations \eqref{eq:Rdot}, \eqref{eq:E}, and \eqref{eq:thetadot} are sufficient to write a system of ordinary differential equations (ODEs) in $R(t)$, $u(t)$, and $\theta_j(t)$. A standard adiabatic blast wave maintains constant $E$, while energy injection and refreshed shock models may add dependence on $t$ or $u$. The general form of the shock evolution equations are then:
\begin{align}
	\dot{R} &= \frac{4 u \gamma}{4 u^2 +3}c  \label{eq:ODE1}\\
	\dot{\theta}_j &= \frac{1}{2\gamma}\sqrt{\frac{2u^2+3}{4u^2+3}} \frac{\dot{R}}{R}  \label{eq:ODE2} \\
	\dot{u} &= -\frac{E_{sw}(4u^2+3)\beta^2 (3\dot{R}/R + \cot (\theta_j/2) \dot{\theta}_j) - \partial_t E}{2 E_{sw} u (4u^4 + 8u^2 + 3)\gamma^{-4} + \beta \Mej c^2 - \partial_u E}  \label{eq:ODE3} \\
	E_{sw} &\equiv \frac{4\pi}{9} \rho_0 c^2 R^3 f_\Omega\ . \nonumber
\end{align}
Equations \eqref{eq:ODE1}-\eqref{eq:ODE3} define a three dimensional system of ODEs in the variables $(R, u, \theta_j)$ which may be solved numerically or, in certain limits, analytically from appropriate initial conditions.  Once the shock evolution ($R(t)$, $u(t)$, $\theta_j(t)$) is known, the flux is given by the integral:
\begin{align}
	F_\nu(\tobs, \nuobs) &= \frac{1+z}{4\pi \dL^2} \int \! \dd \Omega\ R^2\ \Delta R\  \delta^2\ \epsilon'_{\nu'} \ , \label{eq:flux2}
\end{align}
where $\Delta R$ is the effective shock width contributing to the emission \citep{van-Eerten:2010aa}:
\begin{equation}
	\Delta R = \frac{1}{1-\mu \dot{R}} \frac{R}{12\gamma^2} \ . \label{eq:dr}
\end{equation}
 The integrand is evaluated at a constant observer time $\tobs$ and observer frequency $\nuobs$, related to $t$ and $\nu'$ by:
\begin{align}
	\tobs &= (1+z) \left(t - \mu(\theta, \phi) R(t)/c\right)\ , \label{eq:tobs} \\
	\nuobs &= (1+z)^{-1} \delta\ \nu' , \label{eq:nuobs}
\end{align}

@@

\subsection{The Single Shell Approximation -- Structured Case} \label{subsec:algo2}

{\bf
We model a structured jet as a series of concentric top hat jets, each independently obeying the equations of Section \ref{subsec:algo}.  In the continuous case, $R(t)$ and $u(t)$ are promoted to functions of $\theta$ as well: $R(t; \theta)$ and $u(t; \theta)$.  These functions evolve according to their local isotropic-equivalent energy $E(\theta)$.   At the same lab time $t$ (or equivalently observer time $\tobs$) there will be a gradient of both energy $E$ and four-velocity $u$ over the surface of the jet.

For numerical calculations, we approximate the structured jet with a discrete set of $N_\theta$ top-hat jets.  The initial widths of each jet are $\theta_{0, i} = i \thW / N_\theta$ for $i = 1, \dots, N_\theta$.  Each jet is initialized with energy $E_i = E((\theta_{0, i-1} + \theta_{0,i})/2)$ and evolves with independent radius $R_i(t)$, four-velocity $u_i(t)$, and opening angle $\theta_{j, i}(t)$.  The final flux is found by summing the contributions from each top-hat $i$, integrated as in Equation \eqref{eq:flux2}, from the outer edge of jet $i-1$ to its own edge $\theta_{j, i}$.

\begin{align}
		&F_\nu(\tobs, \nuobs) = \frac{1+z}{4\pi \dL^2}  \label{eq:flux3} \\
		&\qquad \times \sum_{i=1}^{N_\theta}  \Bigg [ \int_0^{2\pi} \!\!\! \dd \phi   \int_{\theta_{j, i-1}}^{\theta_{j, i}} \!\!\! \dd \theta \ \sin \theta\  R_i^2\ \Delta R_i\  \delta_i^2\ \epsilon'_{\nu', i} \Bigg ] \nonumber
\end{align}

The integrand of Equation \eqref{eq:flux3} is evaluated at a constant observer time $\tobs$ and hence is a function of both $\phi$ and $\theta$.  Due to jet spreading the integration bounds $\theta_{j, i-1}$ and $\theta_{j, i}$ are also time dependent and hence functions of $\phi$ when evaluated at constant $\tobs$. Equation \eqref{eq:flux3} serves as the basis for all structured jet calculations in \afterglowpy{}.
}

@@

The numerical accuracy of a top-hat light curve with this integration scheme is typically better than $10^{-4}$. {\bf The structured jet calculation splits the integration domain into $N_\theta$ disjoint annuli, each evolved as an independent top-hat and their emission summed following Equation \eqref{eq:flux3}.  The resolution $N_\theta$ is controlled by the {\tt latRes} parameter which sets the number of zones per $\thC$-sized interval: $N_\theta = \mathtt{latRes} \times \thW/\thC$. The choice of  {\tt latRes} has the largest impact on the code performance and accuracy.  We find the default choice $\mathtt{latRes} = 5$ gives sufficiently quick performance at acceptable errors, typically on the order of $10^{-2}$. }

@@

Due to relativistic beaming the flux at any given time is dominated by a small patch of blast wave surface.  The angular coordinates of this patch are $\theta = \theta_*$ and $\phi=0$, it has angular size $\Delta \Omega$, Lorentz factor $\gamma_*$, and emits a specific intensity $I_\nu^*$. We can write the received flux as:
\begin{equation}
	F_\nu \approx I_\nu^*\ \Delta \Omega\ .
\end{equation}

@@

\emph{Structured} --- If the viewer lies outside the core of the blast wave, $\thobs > \thC$, non-trivial angular structure can have a significant effect on the light curve.  {\bf The centroid of emission is not along the direct line of sight ($\theta_* \neq \thobs$) nor on the jet axis ($\theta_* \neq 0$).}  {\bf Material from the energetic regions progressively closer to the jet axis continuously comes into view, becoming the new dominant source of emission, and moving the centroid with time: $0 < \theta_*(\tobs) < \thobs$  with $\theta_* \to 0$ as $\tobs$ increases. The location of the centroid obeys $\gamma^{-1}(\theta_*) \sim |\thobs-\theta_*|$.

The shape of the emitting region $\Delta \Omega$ is complex.  Material closer to the jet core (at $\theta < \theta_*$) has too high a Lorentz factor and is beamed away form the observer.  Material closer to the line of sight (at $\theta > \theta_*$) has a lower Lorentz factor than the material at the centroid, the observer is in its beaming cone but the intensity is dimmer than the centroid material.  This gives the visible patch an asymmetric shape.  The angular size of the visible patch in the azimuthal direction is still controlled by beaming (is of size $\gamma_*^{-1}$), but in the polar direction it is controlled by the steep gradient in $E(\theta)$ instead, resulting in $\Delta \Omega \propto \gamma_*^{-1}$ and $\som \approx 1$.}  The temporal evolution of the light curve depends on both the specific angular structure $E(\theta)$ and the viewing angle $\thobs$. The specific closure relations are given in Table \ref{tab:slopes} and discussed in Section \ref{subsec:structure}.

@@

{\bf For example, our constraints on $\thobs$ and $\thC$ in the fit to the \grbbns{} afterglow are aided immensely by the included gravitational wave constraints on $\thobs$ \citep{Abbott:2017aa}. The afterglow data primarily constrains only the ratio $\thobs/\thC$, leaving $\thobs$ and $\thC$ themselves uncertain and dependent on other parameters like the energy $E_0$ and density $n_0$.  The gravitational wave posterior provides an independent measure of $\thobs$. This leads directly to a constraint on $\thC$ given the tight measure of $\thobs/\thC$ from the afterglow data. }
